---
title: "United States Natality Models 2003-2014"
author: "Daniel Dittenhafer & Justin Hink"
date: "May 22, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
    fig_caption: true
subtitle: "DATA 621: Business Analytics and Data Mining"
---

```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R"), results='hide',message=FALSE}
# Initialize our Rmd environment
```

```{r raw-code, message=FALSE, warning=FALSE, echo=FALSE, code=readLines("../Project/Project-RawCode-dwd.R"), results='hide'}
# Bring in our functions
```

```{r bib-data, echo=FALSE}
bibNatality0306 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Centers for Disease Control and Prevention")),
                 publisher="United States Department of Health and Human Services (US DHHS)",
                 title="Natality public-use data on CDC WONDER Online Database for years 2003-2006 available March 2009",
                 year=2009,
                 note="Accessed: March 1, 2016",
                 url="http://wonder.cdc.gov/natality-v2006.html")

bibNatality0714 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Centers for Disease Control and Prevention")),
                 publisher="United States Department of Health and Human Services (US DHHS)",
                 title="Natality public-use data on CDC WONDER Online Database for years 2007-2014 available February 2016",
                 year=2016,
                 note="Accessed: March 1, 2016",
                 url="http://wonder.cdc.gov/natality-current.html")

bibCensus2015 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="Monthly Postcensal Resident Population, by single year of age, sex, race, and Hispanic origin",
                 year=2015,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/national/asrh/2014/2014-nat-res.html")

bibCensus2000 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="National Intercensal Estimates (2000-2010)",
                 year=2010,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/intercensal/national/nat2010.html")

bibEarnings <- bibentry(bibtype="Misc",
                 author=personList(person(family="Bureau of Labor Statistics")),
                 title="Median wkly earnings, Emp FT, Wage & sal wrkrs, Women - LEU0252882700",
                 year=2015,
                 note="Accessed: March 10, 2016",
                 url="http://data.bls.gov/cgi-bin/surveymost?le")

bibUnemploymentRate <- bibentry(bibtype="Misc",
                 author=personList(person(family="Bureau of Labor Statistics")),
                 title="Labor Force Statistics from the Current Population Survey 2003-2015 - LNS14000000",
                 year=2015,
                 note="Accessed: April 24, 2016",
                 url="http://data.bls.gov/timeseries/LNS14000000")

bibDittenhafer <- bibentry(bibtype="Misc",
                 author=personList(person(family="Dittenhafer", given="Daniel W")),
                 title="U.S. Births & Unemployment Rate 2007 - 2012",
                 year=2014,
                 url="https://github.com/dwdii/DataAcqMgmt/raw/master/FinalProject/FinalProject-Dittenhafer-USBirthsAnalysis.pdf")


bibSpMorg21 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Morgan", given="S. Philip"), 
                                   person(family="Taylor", given="Miles G.")),
                 journal="Annual Review of Sociology",
                 title="Low Fertility at the Turn of the Twenty-First Century",
                 year=2006,
                 month="August",
                 url="http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2849172/")

bibRelWage <- bibentry(bibtype="Article",
                 title="Relative Wage Changes and Fertility in the US",
                 author=person(family="Amialchuk", given="Aliaksandr"),
                 journal="Eastern Economic Journal",
                 year=2013,
                 doi="10.1057/eej.2013.2",
                 volume="39(2)")

bibGitHub <- bibentry(bibtype="Misc",
                      title="NatalityModels",
                      author=personList(person(family="Dittenhafer", given="Daniel W"), 
                                   person(family="Hink", given="Justin")),
                      year=2016,
                      url="https://github.com/dwdii/NatalityModels")

bibGregory <- bibentry(bibtype="Article",
                       title="A Simultaneous Equation Model of Birth Rates in the United States",
                 author=personList(person(family="Gregory", given="Paul R."),
                                   person(family="Campbell", given="John M."),
                                   person(family="Cheng", given="Benjamin S.")),
                 journal="The Review of Economics and Statistics",
                 publisher="The MIT Press",
                 year=1972,
                 doi="10.2307/1924564",
                 url="http://www.jstor.org/stable/1924564",
                 volume="54")

bibMASS <- bibentry(bibtype="Misc",
                    title="Modern Applied Statistics with S",
                    author=personList(person(family="Venables", given="W. N."),
                                      person(family="Ripley", given="B. D.")),
                    publisher="Springer",
                    edition="fourth",
                    address="New York",
                    year=2002,
                    url = "http://www.stats.ox.ac.uk/pub/MASS4",
                    note="ISBN 0-387-95457-0")

bibGgplot <- bibentry(bibtype="Misc",
                      author = person(family="Wickkham", given="Hadley"),
                      title = "ggplot2: Elegant Graphics for Data Analysis",
                      publisher = "Springer-Verlag New York",
                      year = 2009,
                      isbn = "978-0-387-98140-6",
                      url = "http://ggplot2.org")

bibPlyr <- bibentry(bibtype="Misc", 
    title = "The Split-Apply-Combine Strategy for Data Analysis",
    author = person(family="Wickkham", given="Hadley"),
    journal = "Journal of Statistical Software",
    year = 2011,
    volume = 40,
    number = 1,
    pages = "1--29",
    url = "http://www.jstatsoft.org/v40/i01/")

```

# Abstract

The drivers of conception decisions have been studied for some time by economists and sociologists. Consultancies have been established to forecast birth activity, but a publicly available, unified model to forecast births in the United States would aid business and government with planning and resource allocation. In this study we utilize publicly accessible data sets from various United States federal agencies to research relationships and possible predictors of birth counts. The unemployment rate, women’s earnings, gender ratio and census estimates of female population across several age ranges between 2003 and 2014 were considered in the model building process. Both multiple linear regression and generalized linear models were developed and analyzed.  A multiple linear regression model was selected as the best performer, yielding a \(R^2 = `r round(smLmSigLimVars$adj.r.squared, 5)`\). The analysis was performed using the open source R runtime, and the project materials have been shared to a GitHub repository to facilitate reproducible research.

# Keywords

natality, births, demographic theory

# Literature Review

As a starting point for this study, Daniel Dittenhafer has done prior work analyzing births and unemployment rate in the United States. Dittenhafer found a negative relationship between births and unemployment during the time period studied of 2007 - 2012 `r citep(bibDittenhafer)`. Dittenhafer's single predictor linear model using unemployment rate alone yielded an adjusted \(R^2\) of 0.296 with a p-value approaching 0. Although the unemployment-based model showed some interesting analysis, its usefulness is still to be determined. On the other hand, the negative relationship finding appears to be a relevant output of the study, and one which our current research supports. Interestingly, a paper by Gregory, et al from 1972 also studied unemployment rate as an explanatory variable of birth rates, but found it to not be significant `r citep(bibGregory)`.

Morgan and Taylor published a paper in The Annual Review of Sociology regarding recent fertility trends, and specifically a shift to lower birth rates as compared to the second half of the twentieth century `r citep(bibSpMorg21)`. Our research did not relate to this directly, but further reductions in births were observed beginning in 2008. This may constitute another change point or may be a continuation of the trend studied by Morgan and Taylor.

We include women's earnings as a possible predictor in this study based on research by Aliaksandr Amialchuk regarding wage related effects on fertility `r citep(bibRelWage)`. Across all women, women's education, men's education, men's earnings and metro area were all found to be significant in age-specific fertility regression. We were not able to include age specific earnings in our current study, but rather have a single earnings measure for women of child bearing age `r citep(bibEarnings)`.  On a related note, Gregory et al used female labor force participation rate as an indicator of the cost of a mother's time. More participation would suggest lower births in their model. 

# Methodology

## Technology

We used the R runtime via RStudio as our primary environment for all data steps including data preparation, exploration, analysis, model development, validation and selection.  The primary reason for this was the large number of libraries available for both the data cleaning and analysis parts of the project `r citep(bibMASS)`, `r citep(bibGgplot)` `r citep(bibPlyr)`. The work should be portable to other data anlaysis languages (such as Python/numpy/pandas) without a large amount of effort.

A GitHub repository was setup to facilitate collaboration amoungst the team, as well as to share our work in the spirit of reproducible research `r citep(bibGitHub)`.

## Data Preparation

Data sets from the Census Bureau, Centers for Disease Control, and Bureau of Labor Statistics were identified and downloaded to our project GitHub repository. These data sets were programatically joined together in order to provide a unified data set for analysis and modeling. 

### Natality Data

The natality data, which includes birth counts per month, was acquired from the Centers for Disease Control and Prevention in two data sets. The first data set contains data for the years 2003 - 2006 `r citep(bibNatality0306)`. The second data set contains data for the years 2007 - 2014 `r citep(bibNatality0714)`. The data sets were merged together and augmented with additional census, earning and unemployment data as described in the following sections. 

### Census Data

For the period of May 2010 - Decemeber 2015, the Census Bureau's census data was available as monthly population estimates broken down by age and gender `r citep(bibCensus2015)`. The age data was in whole year granulatity (i.e. 5 females age 10, 4 females age 11). We created 3 buckets (10 years each) for the female population by age: 15-24, 25-34, and 35-44.

For the period of 2000 - April 2010, monthly population estimates were only available for the total population `r citep(bibCensus2000)`. We used annual age and gender estimates from the Census Bureau's 2000 - 2010 time period (converted to ratios) to divide the monthly total population into age and gender bins as shown in the following expressions:

**Gender Bins**

For each year, 2003 - 2010:

\[ G_{year} = \frac{F_{year}}{P_{year}} \]

\[ F_{month} = P_{month} * G_{year} \]

\[ M_{month} = P_{month} - F_{month}  \]

Where:

\(\quad G \quad \text{Gender Ratio}\)

\(\quad F \quad \text{Total females, TOT\_FEMALE} \)

\(\quad M \quad \text{Total males, TOT\_MALE} \)

\(\quad P \quad \text{Total population, TOT\_POP} \)


**Age Bins**

Again, for each year, 2003 - 2010:

\[F_{year\_x\_y} = \sum_{i=x}^{y-1} F_{year\_i}\]

\[ A_{year\_x\_y} = \frac{F_{year\_x\_y}}{F_{year}}  \]

\[ F_{month\_x\_y} = F_{month} * A_{year\_x\_y}  \]

Where:

\(\quad x \quad \text{Lower age bound of bin} \)

\(\quad y \quad \text{Upper age bound of bin} \)

\(\quad A \quad \text{Age bin's ratio} \)

### Earnings Data

The earnings data was acquired from the Bureau of Labor Statistics and specifically covers women's weekly earnings from 2003 - 2015 `r citep(bibEarnings)`. The acquired data was at a quarter year granularity and was transformed to a monthly granularity for use in this study by simply assigning a quarter's weekly earnings to each of the related 3 months in the 12 month annual period.

### Unemployment Data

Unemployment data (U3) was acquired from the Bureau of Labor Statistics. The data was at a monthly granularity with no transformations applied before use in the study `r citep(bibUnemploymentRate)`. 

## Data Exploration

We conducted exploratory data analysis to better understand the relationships in the data including correlations, feature distributions and basic summary statistics. 

## Model Development

Ten models were developed and examined for significance using a training subset (80%) of the original full data. Gaussian, Poisson and Negative Binomial linear models were fit using a variety of predictor variables and their significance, VIFs, adjusted \(R^2\) and Akaike information criteria (AIC) were examined.

## Model Validation

A validation data set (VS) was created from a subset of the full data set (20%). This VS data set was used to test how well our candidate models generalize to unseen data. The validation metric for the linear models is the mean squared error from the validation set. 

# Results 

## Data Exploration

### Correlations

The following table shows the correlation coefficients associated with each variable and the dependent variable, *Births*.

```{r, echo=FALSE, tab.cap="Pearson's r Correlation Coefficients"}
kable(corMatrix[order(corMatrix[,"Births"], decreasing=TRUE), c("Births")])
```

### Seasonality

As one might expect, seasonality is quite apparent in the birth data. As shown in the scatter plot, below, August is a very popular month for births. July and September are close behind. This suggests that many conceptions are occuring during the United States holiday season between Thanksgiving and New Years.

```{r month-viz, echo=FALSE, warning=FALSE, fig.height=3, fig.width=6.5, fig.cap="Month vs Births"}
v1 <- exploreVar(allData, "Month", "Births")
v1[[3]]
```

### Gender Ratio

The computed gender ratio which was used to enable the gender buckets for the period of 2003 - 2010 can be seen in the scatterplot below. For these years, the gender ratio is constant for all months of a given year while the birth counts fluctuate. Interestingly, the proportion of females has been dropping steadily, though only slightly during the time period in this study.

```{r genderratio-viz, echo=FALSE, warning=FALSE, fig.height=2.5, fig.cap="GenderRatio vs Date, GenderRatio vs Births"}
v1 <- exploreVar(allData, "GenderRatio", "Births", xunit="% female", rotateLabels=FALSE)
grid.arrange(gGenderRatio, v1[[3]], ncol=2)
```

### Earnings

Women's weekly earnings as a broad median value, as reported by the Current Population Survey via the Bureau of Labor Statistics, revealed a negative correlation with births, as previously shown. 

```{r earnings-viz, echo=FALSE, warning=FALSE, fig.height=2.5, fig.cap="Earnings vs Births"}
v1 <- exploreVar(allData, "Earnings", "Births", xunit="$")
v1[[3]]
```

As part of variance inflation factor (VIF) analyses, we found that the earnings measure we used was correlated with female population levels. As shown, in some cases this was a positive relationship (both 15-24 and 25-34), but for the 35-44 age range this was a negative relationship. In general, these relationships resulted in VIFs which significantly exceeded our threshold of 10 when earnings and the female age values were included in a model. Further study on this relationship may be warranted in order to better understand the drivers behind this effect.

```{r earnings-fm15-fm25-viz, echo=FALSE, warning=FALSE, fig.height=2.5, fig.cap="Earnings vs Female Population 25-34 and 35-44"}
vF15 <- exploreVar(allData, "Earnings", "FEMALE_15_24", xunit="$")
vF25 <- exploreVar(allData, "Earnings", "FEMALE_25_34", xunit="$")
vF35 <- exploreVar(allData, "Earnings", "FEMALE_35_44", xunit="$")
grid.arrange(vF25[[3]], vF35[[3]], ncol=2)
```

## Model Development

We started with all possible predictor variables from our data sets in a multiple linear regression model. This model yielded an adjusted \(R^2\) of `r round(smLmAllVars$adj.r.squared, 5)`, but had significant variance inflation factor issues (> 120,000,000). 

Next, we took the significant predictors from the "All Variables" model and created a new, smaller model. Again, VIFs were large. Specifically, the population bins were highly related.

```{r, vif-sigvars, echo=FALSE, tab.cap="Signficant Variables Linear Model VIFs"}
kable(vfSigVars)
```

A new model using guidance from the correlation analysis yielded our "High Correlation Variables" linear model. VIFs were better, but still well above our threshold of 10. Our fourth model was created using R's *step* function. The "Step" model had an impressive adjusted \(R^2\) of `r round(smLmStep$adj.r.squared, 5)` for only 5 predictor variables (all significant at \(\alpha < 0.05\)). The variance inflation factors were improved, but unfortunately, still exceeded our threshold.

Next, we experimented with a revised "Significant Variables Minus" model which was based on the "Significant Variables" model but with three high VIF variables removed. Again, we saw high VIFs. Around this time, the relationship between the female age bins and earnings, and its impact on our data analysis became apparent.

A review of the variables and the introduction of a lag variable, \(Month9Ago = Month - 9\), brought us to the next model, "Significant Limited". Again a multiple linear regression model, the "Significant Limited" model included 4 significant variables with an adjusted \(R^2 = `r round(smLmSigLimVars$adj.r.squared, 5)`\) and no VIF issues. This was promising, but we continued our investigation of possible models.

```{r modelLM-siglimited-coef, echo=FALSE, tab.cap="Significant Variables Limited Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmSigLimVars))
```

\pagebreak 

```{r, vif-siglimited, echo=FALSE, tab.cap="Significant Variables Limited Linear Model VIFs"}
kable(vfSigLimVars)
```

A Poisson generalized linear model version of the "Significant Limited"" model was produced next. Like its sister model, the "Poisson Significant Ltd" looked good.

```{r modelPM-pmsiglim-coef, echo=FALSE, tab.cap="Poisson Significant Limited Model Coefficient Estimates"}
kable(coefficientsPrep(smPmSigLimVars))
```

A component we felt was missing was interaction between population and month. We again developed a model based on the "Signifcant Limited" model, but this time included an interaction term for *FEMALE_25_34* and *Month9Ago*. This did not perform as we anticipated and instead introduced VIF issues without any predictive benefit.

For our final two models we used R's *stepAIC* to produce Poission and Negative Binomial generalized linear models, respectively.  The "Negative Binomial Step" looked promising also with 6 predictor variables, all but one significant, and no VIF issues.

## Model Validation

As mentioned earlier, a validation data set was reserved for use confirming the performance of each of the developed models. We ran each of the ten previously described models through the validation data and computed the mean squared error (MSE) of the resulting model output. We also captured the Akaike Information Criterion (AIC) for each of the models for reference.

The results of the model validation are shown below. 

```{r table-lm-cverror, echo=FALSE, tab.cap="Linear Model Validation Error Results"}
cvLmResults <- cvLmResults[order(cvLmResults$Val.Error, decreasing=FALSE),]
colnames(cvLmResults) <- c("Model", "VS Error", "Adj R^2", "AIC", "Variables", "VIF")
rownames(cvLmResults) <- seq(1, nrow(cvLmResults))
kable(cvLmResults)
```

The variance inflation factor issues limited us to the "Significant Limited" and "Poisson Significant Ltd" models for further investigation. These models were the least complex and were in the median range in terms of validation performance.  One unexepected negative was that the  AIC value for the "Poisson Significant Ltd" model was higher than anticipated.

### In Depth: Significant Limited Linear Model

The Significant Limited model has an F-statistic of `r round(smLmSigLimVars$fstatistic[1], 2)` and a mean squared error (MSE) of `r round(mse(smLmSigLimVars),2)`.

\[
  \begin{array}{llll}
      y_{births} = & `r lmSigLimVars$coefficients[1]` & +`r lmSigLimVars$coefficients[2]` x_{Month}  &   \\
                   &  & + `r lmSigLimVars$coefficients[3]` x_{Month9Ago}  & \\
                   &  & `r lmSigLimVars$coefficients[4]` x_{FEMALE\_25\_34}  & \\
                   &  & `r lmSigLimVars$coefficients[5]` x_{UnemploymentRate} &  
  \end{array}
\]

We can interpret the coefficients in the following manner. Holding all other predictors constant, for variable:

* *Month*, as the month of the year increased,  a \(`r round(abs(smLmSigLimVars$coefficients[2]), 2)`\) increase in births would occur.
* *Month9Ago*, as the 9 month lagged month of the year increased, a \(`r round(abs(smLmSigLimVars$coefficients[3]), 2)`\) increase in births would occur.
* *FEMALE_25_34*, a unit increase in the population of females age 25-34 would yield a \(`r round(abs(smLmSigLimVars$coefficients[4]), 2)`\) decrease in births.
* *UnemploymentRate*, a unit increase in the *UnemploymentRate* related to a \(`r round(abs(smLmSigLimVars$coefficients[5]), 2)`\) decrease in births.


Linear regression diagnostic plots are shown below. Residuals appear to be normally distributed and variance seems to be fairly constant. 

```{r diagplots-sigcor, fig.height=5.5, echo=FALSE}
#qqnorm(lmStep$residuals)
#qqline(lmStep$residuals)  # adds diagonal line to the normal prob plot
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(lmSigLimVars)
```

Looking at the inverse response plot, there appears to be a strong linear pattern to the predicted response versus actual. This is a desirable relationship.

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(data.frame(y=modelData$Births, y.hat=predict(lmSigLimVars))) +
  geom_point(aes(x=y, y=y.hat)) +
  myTheme +
  labs(title="Inverse Response Plot")
g1
```

Running an auto-correlation analysis with R's *acf* function shows a possible auto-correlation issue with lag 2 and 6. 

```{r autocor-sigcor, echo=FALSE, fig.height=3}
acf(lmSigLimVars$residuals, main="Significant Limited Linear Model Auto-correlation Plot")
```



### In Depth: Poisson Significant Limited Model

The mathematical form of the Poisson Significant Limited Model is as follows:

\[
  \begin{array}{llll}
       log\left(E(y_{births} | x)\right) = & `r pmSigLimVars$coefficients[1]` & +`r pmSigLimVars$coefficients[2]` x_{Month}  &   \\
                   &  & + `r pmSigLimVars$coefficients[3]` x_{Month9Ago}  & \\
                   &  & `r pmSigLimVars$coefficients[4]` x_{FEMALE\_25\_34}  & \\
                   &  & `r pmSigLimVars$coefficients[5]` x_{UnemploymentRate} &  
  \end{array}
\]

We can interpret the coefficients in the following manner. Holding all other predictors constant, for variable:

* *Month*, as the month of the year increased,  a \(e^{`r pmSigLimVars$coefficients[2]`} = `r exp(pmSigLimVars$coefficients[2])`\) times increase in births would occur.
* *Month9Ago*, as the 9 month lagged month of the year increased, a \(e^{`r pmSigLimVars$coefficients[3]`} = `r exp(pmSigLimVars$coefficients[3])`\) times increase in births would occur.
* *FEMALE_25_34*, a unit increase in the population of females age 25-34 would yield a \(e^{`r pmSigLimVars$coefficients[4]`} = `r exp(pmSigLimVars$coefficients[4])`\) times decrease in births.
* *UnemploymentRate*, a unit increase in the *UnemploymentRate* related to a \(e^{`r pmSigLimVars$coefficients[5]`} = `r exp(pmSigLimVars$coefficients[5])`\) times decrease in births.


Regression diagnostic plots are shown below. Residuals appear to be normally distributed and variance seems to be fairly constant. The Leverage plot in the lower right shows many points which exceed Cook's distance and suggest points of high leverage.

```{r diagplots-psigltd, fig.height=5.5, echo=FALSE}
#qqnorm(lmStep$residuals)
#qqline(lmStep$residuals)  # adds diagonal line to the normal prob plot
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(pmSigLimVars)
```

Looking at the inverse response plot, there again appears to be a very visible linear pattern to the predicted response versus actual.

```{r, inverse-psigltd, echo=FALSE, fig.height=3}
g1 <- ggplot(data.frame(y=modelData$Births, y.hat=predict(pmSigLimVars, type="response"))) +
  geom_point(aes(x=y, y=y.hat)) +
  myTheme +
  labs(title="Inverse Response Plot")
g1
```

Running the auto-correlation analysis with R's *acf* function shows the same possible auto-correlation issue with lag 2 and 6 that we saw with the Significant Limited Linear Model.

```{r autocor-psigltd, echo=FALSE, fig.height=3}
acf(pmSigLimVars$residuals, main="Significant Limited Poisson Model Auto-correlation Plot")
```


# Summary

Given the data set inputs we chose for this study, the Significant Limited multiple linear regression model offers the best outcome of those we analyzed. With that said, it has limitations. 

The model does not account for population growth sufficiently, but instead reduces births as more females are present in the age range of 25 - 34. On the otherhand, the unemployment rate coefficient indicates that as more people are unemployed, less pregnancies will result. For us, this is an intuitive result. Another important point is that there appears to be a change in birth activity beginning during the 2007/2008 recession. This change may affect the relationship between the predictor variables and the dependent variable, *Births*.

In terms of future work, we feel the earnings data set was too broad in the sense that it applied to women at all education levels and one large age range. Focusing this data set, possibly including multiple variables across various education levels and age ranges, would be an area which could yield improvement. Additionally, determining an appropriate factor for capturing population growth seems like an important next step. Our model appears to capture the United States holiday season between Thanksgiving and New Year's Day fairly well. Generalizing this somehow for culture agnostic modeling is another area for possible improvement.


# Appendix: Supplemental Tables & Figures

## Visualization: Model vs Validation Set

The following plots show visually the performance of the two selected models against the validation data set.

```{r, echo=FALSE, fig.height=3}
gPdSigLimVarsCV
```

```{r, echo=FALSE, fig.height=3}
gPdPmSigLimVarsCV
```

* Green = Validation Set
* Pink = Model Prediction

# Appendix: Raw Code

```{r, code=readLines("../Project/NatalityModels-DataLoadFuncs.R"), eval=FALSE, highlight=FALSE}
```

```{r, code=readLines("../Project/Project-RawCode-dwd.R"), eval=FALSE, highlight=FALSE}
```


# References

```{r bib-references, results='asis', echo=FALSE, message=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

