---
title: "Natality Models"
author: "Daniel Dittenhafer & Justin Hink"
date: "April 24, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 621: Business Analytics and Data Mining"
---

```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R"), results='hide',message=FALSE}
# Initialize our Rmd environment
```

```{r bib-data, echo=FALSE}
bibCensus2015 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="Monthly Postcensal Resident Population, by single year of age, sex, race, and Hispanic origin",
                 year=2015,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/national/asrh/2014/2014-nat-res.html")

bibCensus2000 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="National Intercensal Estimates (2000-2010)",
                 year=2010,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/intercensal/national/nat2010.html")

bibEarnings <- bibentry(bibtype="Misc",
                 author=personList(person(family="Bureau of Labor Statistics")),
                 title="Median wkly earnings, Emp FT, Wage & sal wrkrs, Women - LEU0252882700",
                 year=2015,
                 note="Accessed: March 10, 2016",
                 url="http://data.bls.gov/cgi-bin/surveymost?le")

bibUnemploymentRate <- bibentry(bibtype="Misc",
                 author=personList(person(family="Bureau of Labor Statistics")),
                 title="Labor Force Statistics from the Current Population Survey 2003-2015 - LNS14000000",
                 year=2015,
                 note="Accessed: April 24, 2016",
                 url="http://data.bls.gov/timeseries/LNS14000000")

```

# Abstract

# Keywords

Natality, Births

# Literature Review

# Methodology

## Data Preparation

Data sets from the Census Bureau, Centers for Disease Control, and Bureau of Labor Statistics were joined together in order to provide a unified data set for analysis and modeling. 

### Natality Data

### Census Data

For the period of May 2010 - Decemeber 2015, the Census Bureau's census data was available as monthly population estimates broken down by age and gender `r citep(bibCensus2015)`. The age data was in whole year granulatity and we
created 10 year buckets for the female population by age: 15-24, 25-34, and 35-44.

For the period of 2000 - April 2010, monthly population estimates were only available for the total population `r citep(bibCensus2000)`. We used annual age and gender estimates from the Census Bureau 2000 - 2010 time period, converted to ratio's, to divide the monthly total population into age and gender bins as shown in the following expressions:

**Gender Bins**

\[ G_{year} = \frac{F_{year}}{P_{year}} \]

\[ F_{month} = P_{month} * G_{year} \]

\[ M_{month} = P_{month} - F_{month}  \]

Where:

\(\quad G \quad \text{Gender Ratio}\)

\(\quad F \quad \text{Total females, TOT\_FEMALE} \)

\(\quad M \quad \text{Total males, TOT\_MALE} \)

\(\quad P \quad \text{Total population, TOT\_POP} \)


**Age Bins**

\[F_{year\_x\_y} = \sum_{i=x}^{y-1} F_{year\_i}\]

\[ A_{year\_x\_y} = \frac{F_{year\_x\_y}}{F_{year}}  \]

\[ F_{month\_x\_y} = F_{month} * A_{year\_x\_y}  \]

Where:

\(\quad x \quad \text{Lower age bound of bin} \)

\(\quad y \quad \text{Upper age bound of bin} \)

\(\quad A \quad \text{Age bin's ratio} \)

### Earnings Data

The earnings data was acquired from the Bureau of Labor Statistics and specifically covers women's weekly earnings from 2003 - 2015 `r citep(bibEarnings)`. The acquired data was at a quarter year granularity and was transformed to a monthly granularity for use in this study where a given quarter's weekly earnings were assigned to each of the 3 months in the 12 month annual period.

### Unemployment Data

Unemployment data (U3) was acquired from the Bureau of Labor Statistics. The data was at a monthly granularity with no transformations applied before use in the study `r citep(bibUnemploymentRate)`. 

# Results 

# Summary

# Appendix: Supplemental Tables & Figures

# Appendix: Raw Code

# References

```{r bib-references, results='asis', echo=FALSE, message=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

