---
title: "Natality Models"
author: "Daniel Dittenhafer & Justin Hink"
date: "April 24, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 621: Business Analytics and Data Mining"
---

```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R")}
# Initialize our Rmd environment
```

```{r bib-data, echo=FALSE}
bibCensus2015 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="Monthly Postcensal Resident Population, by single year of age, sex, race, and Hispanic origin",
                 year=2015,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/national/asrh/2014/2014-nat-res.html")

bibCensus2000 <- bibentry(bibtype="Misc",
                 author=personList(person(family="Census Bureau")),
                 title="National Intercensal Estimates (2000-2010)",
                 year=2010,
                 note="Accessed: April 24, 2016",
                 url="http://www.census.gov/popest/data/intercensal/national/nat2010.html")


```

# Abstract

# Keywords

# Literature Review

# Methodology

## Data Preparation

Data sets from the Census Bureau, Centers for Disease Control, and Bureau of Labor Statistics were joined together in order to provide a unified data set for analysis and modeling. 

### Census Data

For the period of May 2010 - Decemeber 2015, the Census Bureau's census data was available as monthly population estimates broken down by age and gender `r citep(bibCensus2015)`. For the period of 2000 - April 2010, monthly population esimates were only available for the total population `r citep(bibCensus2000)`. We used annual age and gender estimates from the Census Bureau the 2000 - 2010 time period, converted to ratio's, to divide the monthly total population into age and gender bins.

# Results 

# Summary

# Appendix: Supplemental Tables & Figures

# Appendix: Raw Code

# References

```{r bib-references, results='asis', echo=FALSE, message=FALSE}
BibOptions(style="html", bib.style="authortitle")
bibliography()
```

