---
title: "Natality Models Data Exploration"
author: "Daniel Dittenhafer & Justin Hink"
date: "April 24, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 621: Business Analytics and Data Mining"
---

```{r rmd-init, echo=FALSE, code=readLines("../RmdInit.R")}
# Initialize our Rmd environment
```

```{r raw-code, message=FALSE, warning=FALSE, echo=FALSE, code=readLines("./Project-RawCode-dwd.R")}
# Bring in our functions
```

# Data Exploration

The unified data set for this project contains `r nrowAllData` rows of data with 1 response variable and `r ncolAllData - 1` predictor variables. An exploration of this data follows.

## Missing Values

An analysis of missing values in the data set revealed `r nrow(dfMissingVals[dfMissingVals$Missing > 0,])` variables with incomplete data.

```{r, echo=FALSE, eval=TRUE}
#kable(dfMissingVals)
```

## Correlations

The following table shows Pearson's \(r\) correlation coefficients between the numeric independent variables and the response variable *Births*.

```{r, echo=FALSE, tab.cap="Pearson's r Correlation Coefficients"}
kable(corMatrix[order(corMatrix[,"Births"], decreasing=TRUE), c("Births")])
```


## Variable Month


The *Month* variable is the month of birth. As one should expect, the distribution is uniform, but we can see some seasonality to the relationship between *Births* and *Month* with July and August being high frequency birth months. 


```{r month-stats, echo=FALSE, warning=FALSE, tab.cap="Month Variable Statistics"}
v1 <- exploreVar(allData, "Month", "Births")
kable(v1[[1]])
```

```{r month-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r month-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable TOT_POP

The *TOT_POP* variable is the total population per month as esimated by the Census Bureau. 

```{r totpop-stats, echo=FALSE, warning=FALSE, tab.cap="TOT_POP Variable Statistics"}
v1 <- exploreVar(allData, "TOT_POP", "Births")
kable(v1[[1]])
```

```{r totpop-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r totpop-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

