---
title: "Natality Models Data Exploration"
author: "Daniel Dittenhafer & Justin Hink"
date: "April 24, 2016"
documentclass: article 
geometry: margin=0.75in
output: 
  pdf_document:
    number_sections: yes
subtitle: "DATA 621: Business Analytics and Data Mining"
---

```{r rmd-init, message=FALSE, echo=FALSE, code=readLines("../RmdInit.R")}
# Initialize our Rmd environment
```

```{r raw-code, message=FALSE, warning=FALSE, echo=FALSE, code=readLines("./Project-RawCode-dwd.R")}
# Bring in our functions
```

# Data Exploration

The unified data set for this project contains `r nrowAllData` rows of data with 1 response variable and `r ncolAllData - 1` predictor variables. An exploration of this data follows.

## Missing Values

An analysis of missing values in the data set revealed `r nrow(dfMissingVals[dfMissingVals$Missing > 0,])` variables with incomplete data.

```{r, echo=FALSE, eval=TRUE}
#kable(dfMissingVals)
```

## Correlations

The following table shows Pearson's \(r\) correlation coefficients between the numeric independent variables and the response variable *Births*.

```{r, echo=FALSE, tab.cap="Pearson's r Correlation Coefficients"}
kable(corMatrix[order(corMatrix[,"Births"], decreasing=TRUE), c("Births")])
```


## Variable Month


The *Month* variable is the month of birth. As one should expect, the distribution is uniform, but we can see some seasonality to the relationship between *Births* and *Month* with July and August being high frequency birth months. 


```{r month-stats, echo=FALSE, warning=FALSE, tab.cap="Month Variable Statistics"}
v1 <- exploreVar(allData, "Month", "Births")
kable(v1[[1]])
```

```{r month-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r month-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable TOT_POP

The *TOT_POP* variable is the total population per month as esimated by the Census Bureau. 

```{r totpop-stats, echo=FALSE, warning=FALSE, tab.cap="TOT_POP Variable Statistics"}
v1 <- exploreVar(allData, "TOT_POP", "Births", rotateLabels=TRUE, xunit="1000s")
kable(v1[[1]])
```

```{r totpop-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r totpop-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable TOT_FEMALE

The *TOT_FEMALE* variable is the total population of females per month as estimated by the Census Bureau. 

```{r totfemale-stats, echo=FALSE, warning=FALSE, tab.cap="TOT_FEMALE Variable Statistics"}
v1 <- exploreVar(allData, "TOT_FEMALE", "Births", xunit="1000s")
kable(v1[[1]])
```

```{r totfemale-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r totfemale-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable FEMALE_15_24

The *FEMALE_15_24* variable is the total population of females ages 15-24 per month as estimated by the Census Bureau. 

```{r fem1524-stats, echo=FALSE, warning=FALSE, tab.cap="FEMALE_15_24 Variable Statistics"}
v1 <- exploreVar(allData, "FEMALE_15_24", "Births", xunit="1000s")
kable(v1[[1]])
```

```{r fem1524-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r fem1524-viz2, echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable FEMALE_25_34

The *FEMALE_25_34* variable is the total population of females ages 25-34 per month as estimated by the Census Bureau. 

```{r fem2534-stats, echo=FALSE, warning=FALSE, tab.cap="FEMALE_25_34 Variable Statistics"}
v1 <- exploreVar(allData, "FEMALE_25_34", "Births", xunit="1000s")
kable(v1[[1]])
```

```{r fem2534-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r fem2534-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable FEMALE_35_44

The *FEMALE_35_44* variable is the total population of females ages 35-44 per month as estimated by the Census Bureau. 

```{r fem3544-stats, echo=FALSE, warning=FALSE, tab.cap="FEMALE_35_44 Variable Statistics"}
v1 <- exploreVar(allData, "FEMALE_35_44", "Births", xunit="1000s")
kable(v1[[1]])
```

```{r fem3544-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r fem3544-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```




## Variable TOT_MALE

The *TOT_MALE* variable is the total population of females per month as esimated by the Census Bureau. 

```{r totmale-stats, echo=FALSE, warning=FALSE, tab.cap="TOT_MALE Variable Statistics"}
v1 <- exploreVar(allData, "TOT_MALE", "Births", xunit="1000s")
kable(v1[[1]])
```

```{r totmale-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r totmale-viz2, echo=FALSE, warning=FALSE, fig.height=2.5,, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable GenderRatio

The *GenderRatio* variable is the percentage of the total population which are females per month derived from data from the Census Bureau. In cases where month data was not available, the annual gender ratio was computed and applied to the monthly total population. 

```{r genderratio-stats, echo=FALSE, warning=FALSE, tab.cap="GenderRatio Variable Statistics"}
v1 <- exploreVar(allData, "GenderRatio", "Births", xunit="% female", rotateLabels=TRUE)
kable(v1[[1]])
```

```{r genderratio-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r genderratio-viz2, echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable Earnings

The *Earnings* variable is womoen's weekly earnings in current dollars based on data from the Bureau of Labor Statistics. The original values were provided quarterly and were expanded to a monthly format for data analysis purposes.

```{r earnings-stats, echo=FALSE, warning=FALSE, tab.cap="Earnings Variable Statistics"}
v1 <- exploreVar(allData, "Earnings", "Births", xunit="$")
kable(v1[[1]])
```

```{r earnings-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r earnings-viz2, echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```

## Variable UnemploymentRate

The *UnemploymentRate* variable is the unemployment rate per month (U3) based on data from the Bureau of Labor Statistics.

```{r urate-stats, echo=FALSE, warning=FALSE, tab.cap="UnemploymentRate Variable Statistics"}
v1 <- exploreVar(allData, "UnemploymentRate", "Births", xunit="% U3")
kable(v1[[1]])
```

```{r urate-viz, echo=FALSE, warning=FALSE, fig.height=2.5}
grid.arrange(v1[[2]], v1[[3]], ncol=2)
```

```{r urate-viz2, echo=FALSE, warning=FALSE, fig.height=2.5, fig.width=3}
grid.arrange(v1[[4]], ncol=1)
```


# Build Models

## All Variables Linear Model

The first multiple linear regression model uses all 10 predictor variables. The adjusted \(R^2\) value for this model is `r round(smLmAllVars$adj.r.squared, 5)`.

```{r modelLM-allvars-coef, echo=FALSE, tab.cap="All Variables Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmAllVars))
```

```{r, vif-allvars, echo=FALSE, tab.cap="All Variables Linear Model VIFs"}
kable(vfAllVars)
```

## Signficant Variables Linear Model

The second multiple linear regression model uses predictor variables indicated as significant from the All Variables model. The variables selected here were based on the All Variables model prior to the inclusion of the Month9Ago generated variable which appears to have affected the significant variables. The adjusted \(R^2\) value for this model is `r round(smLmSigVars$adj.r.squared, 5)`.

```{r modelLM-sigvars-coef, echo=FALSE, tab.cap="Signficant Variables Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmSigVars))
```

```{r, vif-sigvars, echo=FALSE, tab.cap="Signficant Variables Linear Model VIFs"}
kable(vfSigVars)
```

## High Correlation Variables Linear Model

The third multiple linear regression model uses the six predictor variables with the highest correlation. The adjusted \(R^2\) value for this model is `r round(smLmHighCorVars$adj.r.squared, 5)`.

```{r modelLM-highcorvars-coef, echo=FALSE, tab.cap="High Correlation Variables Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmHighCorVars))
```

```{r, vif-highcorvars, echo=FALSE, tab.cap="High Correlation Variables Linear Model VIFs"}
kable(vfHighCorVars)
```

## Step Linear Model

The *step* function was used to produce the next multiple linear regression model. The adjusted \(R^2\) value for this model is `r round(smLmStep$adj.r.squared, 5)`.

```{r modelLM-step-coef, echo=FALSE, tab.cap="Step Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmStep))
```

```{r, vif-step, echo=FALSE, tab.cap="Step Linear Model VIFs"}
kable(vfStep)
```

## Significant Variables Minus Linear Model

The next model was aimed at removing variables with multicolinearity evidenced by the high VIFs we'd seen on earlier models. The adjusted \(R^2\) value for this model is `r round(smLmSigVifVars$adj.r.squared, 5)`.

```{r modelLM-sigminus-coef, echo=FALSE, tab.cap="Significant Variables Minus Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmSigVifVars))
```

```{r, vif-sigminus, echo=FALSE, tab.cap="Significant Variables Minus Linear Model VIFs"}
kable(vfSigVifVars)
```

## Significant Variables Limited Linear Model

A manual review of features and the introduction of a 9 month lag variable brought us to the next model. The adjusted \(R^2\) value for this model is `r round(smLmSigLimVars$adj.r.squared, 5)`.

```{r modelLM-siglimited-coef, echo=FALSE, tab.cap="Significant Variables Limited Linear Model Coefficient Estimates"}
kable(coefficientsPrep(smLmSigLimVars))
```

```{r, vif-siglimited, echo=FALSE, tab.cap="Significant Variables Limited Linear Model VIFs"}
kable(vfSigLimVars)
```


## Poisson Significant Limited Model

A Poisson generalized linear model version of the Significant Limited model was produced next. The deviance value for this model is `r round(smPmSigLimVars$deviance, 5)`.

```{r modelPM-pmsiglim-coef, echo=FALSE, tab.cap="Poisson Significant Limited Model Coefficient Estimates"}
kable(coefficientsPrep(smPmSigLimVars))
```

```{r, vif-pmsiglim, echo=FALSE, tab.cap="Poisson Significant Limited Model VIFs"}
kable(vifPmSigLimVars)
```

## Poisson Step Model

The *stepAIC* function was used to produce the next a Poisson generalized linear model. The deviance value for this model is `r round(smPmStepAICSuggested$deviance, 5)`.

```{r modelPM-pmstep-coef, echo=FALSE, tab.cap="Poisson Step Model Coefficient Estimates"}
kable(coefficientsPrep(smPmStepAICSuggested))
```

```{r, vif-pmstep, echo=FALSE, tab.cap="Poisson Step Model VIFs"}
kable(vifPmStep)
```


## Negative Binomial Step Model

The *stepAIC* function was used to produce the next a Negative Binomial generalized linear model. The deviance value for this model is `r round(smNbmStepAIC$deviance, 5)`.

```{r modelNBM-nbmstep-coef, echo=FALSE, tab.cap="Negative Binomial Step Model Coefficient Estimates"}
kable(coefficientsPrep(smNbmStepAIC))
```

```{r, vif-nbmstep, echo=FALSE, tab.cap="Negative Binomial Step Model VIFs"}
kable(vifNbm)
```


# Select Models

A validation data set (VS) was created from a subset of the full dataset for use in the mulitple linear regression. This VS data set was used to perform a level of independent validation of the previously described models. The validation metric for the multiple linear regression models is the mean squared error from the validation set. 

The results of the multiple linear regression model validation are shown below.

```{r table-lm-cverror, echo=FALSE, tab.cap="Linear Model Validation Error Results"}
cvLmResults <- cvLmResults[order(cvLmResults$Val.Error, decreasing=FALSE),]
colnames(cvLmResults) <- c("Model", "VS Error", "Adj R^2", "AIC", "Variables", "VIF")
rownames(cvLmResults) <- seq(1, nrow(cvLmResults))
kable(cvLmResults)
```

Based on the criteria of least complex model with lowest validation error, highest \(R^2\) and/or lowest AIC, and no multicollinearity issues, the Significant Limited model is favored for further investigation.

## Evaluation: Significant Limited Linear Model

The Significant Limited model has an F-statistic of `r round(smLmSigLimVars$fstatistic[1], 2)` and a mean squared error (MSE) of `r round(mse(smLmSigLimVars),2)`.

\[
  \begin{array}{llll}
      y_{births} = & `r lmSigLimVars$coefficients[1]` & +`r lmSigLimVars$coefficients[2]` x_{Month}  &   \\
                   &  & + `r lmSigLimVars$coefficients[3]` x_{Month9Ago}  & \\
                   &  & `r lmSigLimVars$coefficients[4]` x_{FEMALE\_25\_34}  & \\
                   &  & `r lmSigLimVars$coefficients[5]` x_{UnemploymentRate} &  
  \end{array}
\]

We can interpret the coefficients in the following manner. Holding all other predictors constant, for variable:

* *Month*, as the month of the year increased,  a \(`r round(abs(smLmSigLimVars$coefficients[2]), 2)`\) increase in births would occur.
* *Month9Ago*, as the 9 month lagged month of the year increased, a \(`r round(abs(smLmSigLimVars$coefficients[3]), 2)`\) increase in births would occur.
* *FEMALE_25_34*, a unit increase in the population of females age 25-34 would yield a \(`r round(abs(smLmSigLimVars$coefficients[4]), 2)`\) decrease in births.
* *UnemploymentRate*, a unit increase in the *UnemploymentRate* related to a \(`r round(abs(smLmSigLimVars$coefficients[5]), 2)`\) decrease in births.


Linear regression diagnostic plots are shown below. Residuals appear to be normally distributed and variance seems to be fairly constant. There are definitely some outliers which the model is not capturing.

```{r diagplots-sigcor, fig.height=5.5, echo=FALSE}
#qqnorm(lmStep$residuals)
#qqline(lmStep$residuals)  # adds diagonal line to the normal prob plot
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(lmSigLimVars)
```

Looking at the inverse response plot, there does appear to be a good linear pattern to the predicted response versus actual.

```{r, echo=FALSE, fig.height=3}
g1 <- ggplot(data.frame(y=modelData$Births, y.hat=predict(lmSigLimVars))) +
  geom_point(aes(x=y, y=y.hat)) +
  myTheme +
  labs(title="Inverse Response Plot")
g1
```

Running a more targeted auto-correlation analysis with R's *acf* function shows a possible auto-correlation issue with lag 2 and 6.

```{r autocor-sigcor, echo=FALSE, fig.height=3}
acf(lmSigLimVars$residuals, main="Significant Limited Linear Model Auto-correlation Plot")
```

```{r, echo=FALSE}
gPdSigLimVarsCV
```

## Evaluation: Poisson Significant Limited Model

The mathematical form of the Poisson Significant Limited Model is as follows:

\[
  \begin{array}{llll}
       log\left(E(y_{births} | x)\right) = & `r pmSigLimVars$coefficients[1]` & +`r pmSigLimVars$coefficients[2]` x_{Month}  &   \\
                   &  & + `r pmSigLimVars$coefficients[3]` x_{Month9Ago}  & \\
                   &  & `r pmSigLimVars$coefficients[4]` x_{FEMALE\_25\_34}  & \\
                   &  & `r pmSigLimVars$coefficients[5]` x_{UnemploymentRate} &  
  \end{array}
\]

We can interpret the coefficients in the following manner. Holding all other predictors constant, for variable:

* *Month*, as the month of the year increased,  a \(e^{`r pmSigLimVars$coefficients[2]`} = `r exp(pmSigLimVars$coefficients[2])`\) times increase in births would occur.
* *Month9Ago*, as the 9 month lagged month of the year increased, a \(e^{`r pmSigLimVars$coefficients[3]`} = `r exp(pmSigLimVars$coefficients[3])`\) times increase in births would occur.
* *FEMALE_25_34*, a unit increase in the population of females age 25-34 would yield a \(e^{`r pmSigLimVars$coefficients[4]`} = `r exp(pmSigLimVars$coefficients[4])`\) times decrease in births.
* *UnemploymentRate*, a unit increase in the *UnemploymentRate* related to a \(e^{`r pmSigLimVars$coefficients[5]`} = `r exp(pmSigLimVars$coefficients[5])`\) times decrease in births.


Regression diagnostic plots are shown below. Residuals appear to be normally distributed and variance seems to be fairly constant. The Leverage plot in the lower right shows many points which exceed Cook's distance and suggest points of high leverage.

```{r diagplots-psigltd, fig.height=5.5, echo=FALSE}
#qqnorm(lmStep$residuals)
#qqline(lmStep$residuals)  # adds diagonal line to the normal prob plot
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page 
plot(pmSigLimVars)
```

Looking at the inverse response plot, there does appear to be a good linear pattern to the predicted response versus actual.

```{r, inverse-psigltd, echo=FALSE, fig.height=3}
g1 <- ggplot(data.frame(y=modelData$Births, y.hat=predict(pmSigLimVars, type="response"))) +
  geom_point(aes(x=y, y=y.hat)) +
  myTheme +
  labs(title="Inverse Response Plot")
g1
```

Again, running a more targeted auto-correlation analysis with R's *acf* function shows the same possible auto-correlation issue with lag 2 and 6.

```{r autocor-psigltd, echo=FALSE, fig.height=3}
acf(pmSigLimVars$residuals, main="Significant Limited Poisson Model Auto-correlation Plot")
```

```{r, echo=FALSE}
gPdPmSigLimVarsCV
```

## Appendix: Significant Limited vs Full Data Set

The following plot shows the Significant Limited model against the full, combined training and validation data set.

```{r, echo=FALSE}
gPdPmSigLimVarsAll
```